Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-13 paper.tex
modified the eln section to be more concise and focus more on challenges. Still needs some more work

Before the year 2000, many chemists were still using paper
lab-notebooks, multiple compound registration, and search
tools. Compared to today there were many pitfalls and drawbacks
associated with this. The overall architecture was too inflexible to
adopt to fast changing data standards, scientists spent too much time
with administrative work, chemical data quality was inconsistent, an
alignment with other working groups was inefficient (\emph{e.g.}, for
running analytical experiments), a legal dispute required to find data
in paper lab-notebooks, data synchronization between systems caused
lag times and access delays, and local silos hampered an efficient
global collaboration. Now, chemists prefer ELNs, because they can run
searches in the data across the organization. This does not only allow
to find experts faster, but also allows access to highly trusted data
with approved quality workflows. Since this also gives access to
analytical and legal data directly every new ELN release is aiming at
increasing the integration of scientific and business workflows for
reducing the risk due to irrational or bounded-rational decision
making and increasing decision making efficiency.

While ELNs technically existed as long ago as the 1990s, chemists
adopted their use in large numbers in 2004 and 2005, and the market is
still growing~\cite{ELNstatus}. As one report notes, ``The initial
drive for the development of ELNs came from the field of chemistry,
perhaps driven by the early adoption by chemists of computer
technologies \ldots The pharmaceutical industry shifted from a
position of `if we get an ELN' to `when we get an
ELN'.''~\cite{ELNstatus}. The growth of the use of ELNs has been
driven by many factors including ease of use and regulatory
compliance. In chemistry, every experiment requires significant domain
expertise and provenance information (especially in industrial
settings where an experiment might be a key component of a patent
filing). ELNs play an important role in reducing the risks associated
with information loss by allowing easy sharing and retrieval of domain
knowledge, securing sensitive data, ensuring data is time stamped and
in general maintaining an audit trail of all experiemnts and data
involved in a project.

@@

However, the future still leaves many challenges ahead. Using
external/public databases with chemical and bioactivity data remains a
challenge due to differences in ontologies, synchronization and
maintenance, and efficient substructure and similarity search within
complex data types (compounds). A collaboration with external parties,
e.g. contract research, poses other problems with duplicate checking
for compounds (e.g. tautomeric or salt structures) and efficient data
synchronization, and most often a double-registration is required for
fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and thus rely on external services. Cloud services could
help not only to provide a service infrastructure for all parties
involved, but also to provide the required private/public access
toolbox. Furthermore, there are still many data sources not being
indexed properly, since chemistry patents are often cryptic and
image/text mining remains a puzzle. One could argue that the
image/text mining in scientific journals is less challenging; in fact,
it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract
Service) is a highly trusted source, and public chemistry and
bioactivity space has to improve quality and interconnectivity to
compare with it. Efforts like ChEMBL and PubChem-BioAssay are on the
right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring
constant growth, usage, and collaboration scenarios between private
and public parties.  In a business context we need to address ROI
(return-of-investment) questions, not only for single institutions and
companies, but also for collaboration networks of legal partners.  So,
only when the technical and scientific questions are solved can we
ensure that legal questions about contribution metrics will get enough
room to grow. Solving challenges is one thing (science and technical),
ensuring that the contributing partners get rewarded by their
contribution another (legal).

>>>>

Before the year 2000, many chemists were still using paper
lab-notebooks, multiple compound registration, and search
tools. The overall architecture was too inflexible to
adopt to fast changing data standards, scientists spent too much time
with administrative work, chemical data quality was inconsistent, an
alignment with other working groups was inefficient (\emph{e.g.}, for
running analytical experiments). Legal disputes would require manual
searches through many paper lab notebooks and data synchornization was
painful, hampering large scale collaborations.

%  Now, chemists prefer ELNs, because they can run
% searches in the data across the organization. This does not only allow
% to find experts faster, but also allows access to highly trusted data
% with approved quality workflows. Since this also gives access to
% analytical and legal data directly every new ELN release is aiming at
% increasing the integration of scientific and business workflows for
% reducing the risk due to irrational or bounded-rational decision
% making and increasing decision making efficiency.

While ELNs technically existed as long ago as the 1990s, chemists
adopted their use in large numbers in 2004 and 2005, and the market is
still growing~\cite{ELNstatus}. As one report notes, ``The initial
drive for the development of ELNs came from the field of chemistry,
perhaps driven by the early adoption by chemists of computer
technologies \ldots The pharmaceutical industry shifted from a
position of `if we get an ELN' to `when we get an
ELN'.''~\cite{ELNstatus}. The growth of the use of ELNs has been
driven by many factors including ease of use (rapid searches across
organization-wide data sources, easy sharing of domain knowledge) and
regulatory compliance (timestamped experimental data, experiment and
data approvals and QC information). Given that any experiment could
become the core of a patent filing, ELNs allow organizations to define
and implement audit trails in a standardized and efficient fashion.

@@

However, the future still leaves many challenges ahead. Using
external/public databases with chemical and bioactivity data remains a
challenge due to differences in identifiers, synchronization, curation
\& error correctin mechanisms and efficient substructure and
similarity search within complex data types (compounds). A
collaboration with external parties, e.g. contract research, poses
other problems including compound deduplication and efficient data
synchronization. If external partners are small they often do not even
have sufficient IT resources themselves and thus rely on external
services. Cloud services could help not only to provide a service
infrastructure for all parties involved, but also to provide the
required private/public access toolbox. Furthermore, there are still
many data sources not being indexed properly: chemistry patents are
often cryptic and chemical image and text mining remains a challenge
(albeit, currently being addressed in academia and industry
\cite{Jessop:2011fk,Sayle:2009uq}). The closed CAS (Chemical Abstract
Service) is a highly trusted source, and public chemistry and
bioactivity space has to improve quality and interconnectivity to
compare with it. Efforts like ChEMBL and PubChem-BioAssay are on the
right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring
constant growth, usage, and collaboration scenarios between private
and public parties.
